Genetic draft and valley crossing

### 2018-06-04 ms/valley_crossing.tex
edited through Mathematical background

\documentclass[rmp]{revtex4}
\usepackage[english]{babel}
\usepackage{amssymb,amsfonts,amsmath}
\usepackage{color}
\usepackage{breqn}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}

@@

\newcommand{\EQ}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\EQS}[2]{Eqs.~(\ref{eq:#1}) and (\ref{eq:#2})}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\TAB}[1]{Tab.~\ref{tab:#1}}
\newcommand{\REF}[1]{ref.~\citep{#1}}

\newcommand{\comment}[1]{{\color{red}#1}}

@@

\DeclareMathOperator*{\argmin}{arg\,min}

\usepackage{setspace}

@@

Valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
If the number of mutants produced every generation is high, which occurs in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo.  
In small populations, neutral or deleterious intermediate mutants generated by wild type individuals can drift to fixation. Then, the further beneficial mutations necessary for the adaptation can fix due to positive selection.
A third possibility occurs for intermediate population sizes where wild type individuals generate transient mutant subpopulations or ``bubbles".
These bubbles are ordinarily doomed to extinction due to drift, but additional mutations in a lucky bubble may generate a complex adaptation that sweeps to fixation; this process has been referred to as ``tunneling" \citep{iwasa_2004, weissman_2009}.
On its own, valley crossing becomes more likely as the population size increases \citep{weissman_2009}. However, recent work suggests that valley crossing relative to the fixation of a simple beneficial mutation is least likely at intermediate population sizes where tunneling occurs \citet{ochs_2015}. This suggests that tunneling may be the most difficult mode of valley crossing. 
% Valley crossing in sexually reproducing populations was by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
% Fairly low recombination rates can bring mutant subpopulations together, increasing the rate at which complex adaptations arise and sweep.
% High recombination rates, on the other hand, lower the valley crossing rate by forcing fit individuals to outcross with the less fit wild type.

@@

Our model features a population of $N$ haploid individuals in which an effectively infinite number of ``background" loci, with weak fitness effects, are currently segregating.
We focus on two loci, which are initially fixed for alleles $a$ and $b$.
These alleles mutate to $A$ and $B$ at rates $\mu_a$ and $\mu_b$, respectively (we neglect back mutations).
Genotypes $ab$, $Ab$, $aB$, and $AB$ have fitnesses $1$, $1-\delta_a$, $1-\delta_b$, and $1+s$ respectively, with $\delta \geq 0$ (so the intermediate genotypes are deleterious) and $1/N < s < 1$ (so the complex adaptation is strongly beneficial).
For the remainder of our analysis, we will assume that $\mu_a = \mu_b$ and $\delta_a = \delta_b$.
We further assume that the large number of loci of weak effect contribute to a constant background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection, sets the rate $v$ at which the mean fitness advances.
Let $x$ be the background fitness of a lineage.
The distribution of background fitnesses $f(x)$ is assumed to be roughly Gaussian: $f(x) = \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
Our goal is to compute the expected time $\mathbb{E}\left[ T\right]$ for such a complex adaptation to arise and fix in the population, as a function of the parameters.

In very large populations, enough double mutants are generated that the valley crossing time is dominated by the time required for a double mutant to sweep through the population.
At the other extreme, in small populations, the crossing time is dominated by the wait for a deleterious single mutant to fix: only thereafter does a successful double mutant appear.
We begin by zeroing in on the intermediate case: stochastic tunneling, the most interesting and mathematically demanding mode of valley crossing.
Once this is understood, extending our results to incorporate sequential fixation or the arisal of a complex adaptation in a single generation is straightforward.

There are three components to the process of stochastic tunneling.
First, a single mutant lineage must appear, which occurs with rate $N\mu$.
If this lineage arises at time $t_0$ and persists until time $T$, then it gives rise to $W = int_{t_0}^T n(t) dt$ total individuals before going extinct, where $n(t)$ is the number of mutant individuals extant at time $t$.
We refer to such a short-lived mutant lineage, and all the individuals generated within that lineage, as a ``bubble": the time integrated number of individuals $W$ is the ``weight".
During the lifetime of a bubble, the complex adaptation must appear and establish, that is, rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu p_{\mathrm {fix}}$.
Finally, the complex adaptation must sweep to fixation.
The first two steps can be folded in together and considered as one process, so that what is relevant is the expected time until a single mutant lineage arises that is destined to produce a successful double mutant.
In this way, $\mathbb{E}\left[ T \right] = \mathbb{E} \left[ T_0 + T_1 \right]$, where $T_0$ is the time to the first successful bubble--the first that produces a double mutant that is destined for fixation--and $T_1$ is the time required for the double mutant to sweep.
We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term.

Most of the time, a bubble will simply arise and go extinct.
If mutation and fixation are sufficiently rare events, then a double mutant fixing can be modeled as a Poisson process: with probability $1-e^{-\mu p_{\mathrm{fix}} W}$, the bubble will give rise to a successful double mutant.
Computing the expected value $\phi = \left< 1-e^{\mu p_{\mathrm {fix}} W} \right>$ thus gives the probability that a particular bubble will lead to a successful valley crossing.
One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
That is to say, it is not possible to simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves, but rather one must convolute both of them over the background fitness.
Individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
As a result, $W$ will depend on $x$ and $\delta$, and $p_{\mathrm fix}$ will depend on $x$ and $s$.
These will have to be integrated over the fitness distribution $f(x)$ in order to obtain the total crossing probability $\Phi(s,\delta) = \int_{-\infty}^\infty \phi(x,s,\delta) f(x) dx$.
The expected valley crossing time then becomes $1/N\mu\Phi$.

We begin by considering the dynamics of $p_{\mathrm{fix}}(x,s)$ before moving on to consider the distribution of bubble sizes.
We zero in on the ``weak driver" model, in which background mutations are frequent but of weak effect compared to $\sigma$.
In this case, the background beneficial mutation rate $U_b$ and mean fitness effect $\bar{\epsilon}$ turn out not to matter: what is relevant is the product $D = U_b \bar{\epsilon}^2/2$.
Taking the limit $U_b \to \infty$ and $\bar{\epsilon} \to 0$ yields the ``infinitesimal" model.
\citet{hallatschek_2011} and \citet{good_2014} showed that the fixation probability in this regime obeys the differential equation 
\begin{equation}
v\partial_x p(x,s) = x p(x,s) + D\partial_x^2 p(x,s) - \frac{p(x,s)^2}{2},
\end{equation}
in which we have assumed that $v \approx \sigma^2$, which is tantamount to claiming that mutational effects are less important than selection in determining the advance of the mean fitness.
The fixation probability then increases linearly in $x$ above some critical value $x_c$ and remains small otherwise, with $x_c \approx \sigma^4/4D$: when $\sigma$ is large, only mutations in the nose stand any chance of fixing unless they can be dragged to higher fitness by a background mutations (large $D$).

In order to compute $\Phi$ we must next consider the behavior of mutant lineages.
Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm {fix}}$.
\citet{neher_shraiman_2011} showed that $\phi$ follows the equation
\begin{equation}
\partial_t \phi = z + (z + x - \sigma^2 t + \delta)\phi^2 - (1 + x - \sigma^2 t + \delta)\phi^2,
\end{equation}
but their analysis also included a recombination term that made it possible to pull out the integrated Laplace transform $\int_{-\infty}^\infty p(x) \phi(x) dx$ without the need to actually solve this equation or send time to infinity.
Unfortunately this method is not available to us, and the nose dominance of the bubble size distribution, combined with the $x$ dependence of $z$ (via $z = \mu p_{\mathrm{fix}}(x)$), means that a simple closed form expression for $\Phi$ is not forthcoming.
We therefore focus on simulation methods to study the dynamics of valley crossing.

% \iffalse
% (Note: There is an alternate regime in the appendix of \citet{good_2014} that would be better here. Will amend accordingly.)
% We begin by considering the dynamics of $p_{\mathrm {fix}}$ before moving on to consider the distribution of bubble sizes.
% To do so, we follow the method of \citet{good_2012} and \citet{good_2014}.
% Let $p(x,s)$ be the fixation probability of the double mutant as a function of the absolute fitness $x$.
% Furthermore, suppose that individuals experience ``background" mutations of small effect at one of a large number of loci.
% These mutations occur at rate $U$ and are distributed as $\rho(\epsilon)$.
% Then the fixation probability turns out to be governed by the integro-differential equation
% \begin{equation}
% v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
% \end{equation}
% where $X = x-\sigma^2 t+s$ is the fitness relative to the (time dependent) mean fitness.

% We zero in on the ``infinitesimal" model, where mutations are of small effect compared to $\sigma$.
% In that case, the fixation probability is governed by the equation

% \begin{equation}
% v \partial_X p(X) = X p(X) + U \int \left[ p(X+\epsilon) - p(X)\right]\rho(\epsilon) d\epsilon - \frac{p(X)^2}{2},
% \end{equation}

% The perturbation provided by the mutation term is very important: without it, the equation is Bernoulli and can be solved readily, but the only solution is $p(X) = 0$ everywhere.
% This should be intuitive, as lineages in a large population are doomed to extinction as the mean fitness inexorably advances, unless they an increase their fitness via recombination or mutation.
% Assuming that $s$ is small compared to $\sigma$, a matched asymptotics analysis yields
% \begin{equation}
% p(X,s) =
% \begin{cases}
% 0 &\mathrm{~if~} X+s < 0, \\
% X_c e^{\frac{(X+s)^2-X_c^2}{2\sigma^2}} &\mathrm{~if~} 0 < X+s < X_c, \\
% X+s &\mathrm{~if~} X+s > X_c.
% \end{cases}
% \end{equation}
% (If $s$ is large compared to $\sigma$, then the standard Haldane result, $p(s) \sim s$, obtains.)
% Effectively, this means the fixation probability is \emph{zero} if the double mutant lineage's fitness is less than the average for the population: if the fitness is less than some critical value $X_c$, then the lineage must generate a background mutant in order to fix, and if it is above $X_c$, the fixation probability is determined by the lineage's ability to survive drift and establish.
% The value of $X_c$ is determined by the shape of $\rho(\epsilon)$.
% Suppose $\rho$ is distributed exponentially: $\rho(\epsilon) = \frac{1}{\bar{\epsilon}} e^{-\epsilon/\bar{\epsilon}}$.
% Then $X_c$ is determined by solving
% \begin{equation}
% 2 = \frac{\sqrt{2\pi}\sigma U}{\bar{\epsilon}} \left[ \frac{1}{X_c - \frac{\sigma^2}{\bar{\epsilon}}} + \frac{\bar{\epsilon}}{\sigma^2} \left( 1+ \frac{\bar{\epsilon}}{X_c} \right) \right] e^{\frac{(X_c - \sigma^2/\bar{\epsilon})^2}{2\sigma^2}}. 
% \end{equation}

% We now consider the behavior of single mutant bubbles.
% Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu p_{\mathrm {fix}}$.
% We proceed to derive an equation for the time evolution of the Laplace transform.

% Let $p(w,T|k,t,x)$ be the probability that a bubble has reached weight $w$ at time $T$, given that there were $n$ individuals present in the lineage at time $t$, and the bubble has background fitness $x$.
% We have
% \begin{align*}
% -(\partial_t -k\partial_w) p(w,T|k,t,x) = &-k(2+x-\bar{x}-\delta)p(w,T|k,t,x) \\
% & + k(1+x-\bar{x}-\delta)p(w,T|k+1,t,x) \\
% & +kp(w,T|k-1,t,x).
% \end{align*}
% Consider the Laplace transform $\mathcal{L}\left[ p(w,T|k,t,x) \right] = \hat{p}(z,T|k,t,x) =  \int_0^\infty e^{-zw} p(w,T|k,t,x) dw$.
% Using the convolution property of the transform, as well as the fact that each lineage is independent, allows us to write $\hat{p}(w,T|k,t,x) = \hat{p}^k(w,T|1,t,x)$.
% We therefore have
% \begin{equation}
% -\partial_t\hat{p}(z,T|t,x) = -z\hat{p}(z,T|t,x) -(2+x-\bar{x}-\delta)\hat{p}(z,T|t,x) + (1+x-\bar{x}-\delta)\hat{p}(z,T|t,x)^2 + 1.
% \end{equation}
% Defining $\phi = 1-\hat{p}$ (consistent with our previous definition of $\phi$ and normalizing our variables via $\tau = \sigma(T-t)$, $\chi = x/\sigma-\sigma T$, $\tilde{z} = z/\sigma$, and $\tilde{\delta} = \delta/\sigma$ yields
% \begin{equation}
% \partial_\tau \tilde{\phi} = \tilde{z} + (\tilde{z} + \chi + \tau + \tilde{\delta})\tilde{\phi} - (1 + \chi + \tau + \tilde{\delta})\tilde{\phi}^2,
% \end{equation}
% where $\tilde{\phi}$ = $\phi/\sigma$.
% We assume that the prefactor of $\tilde{\phi}^2$ is approximately $1$, which corresponds to the growth rate being small compared to $1$.
% In addition, we will eventually want to take the limit as $\tau \to \infty$, which corresponds to averaging over all possible bubble sizes, no matter how long lived they are.
% However, this will yield the absurd result that $\tilde{\phi}$ grows without bound.
% To remedy this, we assume that the lifespan of bubbles is short compared to the time scale over which selection acts, meaning that we drop the $\tau$ term.
% Defining $\theta = \tilde{z} + \chi + \tilde{\delta}$ yields
% \begin{equation}
% \partial_\tau \tilde{\phi} = \tilde{z} + \theta \tilde{\phi} - \tilde{\phi}^2.
% \end{equation}
% Using matched asymptotics allows us to write
% \begin{equation}
% \tilde{\phi} = 
% \begin{cases}
% \tilde{z} &\mathrm{~if~} \theta < 0, \\
% \theta &\mathrm{~if~} \theta > 0.
% \end{cases}
% \end{equation}
% Intuitively, this corresponds to stating that bubbles with a positive growth rate are likely to persist for some time, but bubbles with a negative growth rate are doomed to extinction: the only way they can succeed is to generate a successful double mutant in $O(1)$ generations.
% Knowing the form of $\phi$ and $z$ allows us to compute the overall tunneling probability and hence the expected time to a tunneling event, once we integrate over the fitness distribution $f(x)$.
% \fi

%\begin{figure}
\begin{figure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/det_tunnel.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
\end{subfigure}
\caption{Valley crossing times in the semi-deterministic tunneling (left) and deterministic fixation (right) regimes. In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^5$ and $\mu = 10^{-4}$, and on the right, $N = 10^6$ and $\mu = 10^{-3}$. Orange lines are averages, boxes are interquartile ranges, and end caps are overall ranges. Each box represents about $100$ simulation runs.}
\label{fig:deterministic}
\end{figure}

\section*{Methods}

@@

@@

Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
By comparing how $\alpha$ changes as a function of the strength of selection on linked variation, $\sigma$, we can determine how $\sigma$ affects the speed of valley crossing relative to simple beneficial sweeps. Higher (or lower) values of $\alpha$ mean that valley crossing times increase (or decrease) relative to the time required for a simple beneficial sweep.
Though the precise value of this ratio depends on $\delta$ and the selection coefficients $s_{\mathrm{valley}}$ of the complex adaptation and $s_{\mathrm{sweep}}$ of the sweeping solo beneficial mutation, a general trend nonetheless emerges.
In this case, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
For several combinations of parameter values, we find that the ratio $\alpha$ is smaller for high $\sigma$ than for low $\sigma$: see figure \ref{fig:ratios}.
In figure \ref{fig:alpha_ratio}, we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
Overall, we find that increasing $\sigma$ increases the extent to which valley crossing is favored \emph{except} when $N\mu > 1$.
In this case, stochastic tunneling ceases to be the dominant mode of valley crossing.
\citet{weissman_2009} referred to this regime as ``semi-deterministic tunneling": a more or less stable number of intermediate mutant individuals $N\mu/\delta$ persists in the population at low frequencies, which give rise to double mutants at a low but constant rate.
At high $\sigma$, the number of intermediate mutants is likely to fluctuate substantially \citep{cvijovic_2017}, but the average behavior is more or less the same.
This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $p_{\mathrm{fix}}$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.

@@

% One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
% That is to say, the total rate of valley crossing will be given by $\int P(x) p_{\mathrm{fix}}(x,s) \mathbb{E}\left[ w(x,\delta) \right] dx$, with $P(x)$ the (presumably Gaussian) background fitness distribution.
% One cannot simply compute $p_{\mathrm{fix}}$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
% This is intuitive: individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
% We proceed to derive both $p_{\mathrm{fix}}(x,s)$ and $P(w)$ in the zero recombination limit.

>>>>

\documentclass[rmp]{revtex4}
\usepackage[english]{babel}
\usepackage{amssymb,amsfonts,amsmath}
\usepackage{xcolor}
\usepackage{breqn}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}

@@

\newcommand{\EQ}[1]{Eq.~\eqref{eq:#1}}
\newcommand{\EQS}[2]{Eqs.~\eqref{eq:#1} and \eqref{eq:#2}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\TAB}[1]{Tab.~\ref{tab:#1}}
\newcommand{\REF}[1]{ref.~\citep{#1}}

\newcommand{\comment}[2][noinline]{\todo[color=red,#1]{#2}}

@@

%%% math macros
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\pfix}{p_{\mathrm{fix}}}

\usepackage{setspace}
\usepackage[colorinlistoftodos]{todonotes}

@@

Valley crossing in asexual populations is increasingly well understood \citep{weissman_2009}.
If the number of mutants produced every generation is high, which occurs in large populations with high mutation rates, a complex adaptation requiring multiple mutations that is destined to fix will eventually be generated de novo. This is sometimes called deterministic fixation of the double mutant \citep{weissman_2009}.
In small populations, neutral or deleterious intermediate mutants generated by wild type individuals can drift to fixation. Then, the further beneficial mutations necessary for the adaptation can fix due to positive selection. This is the sequential fixation regime.
A third possibility occurs for intermediate population sizes where wild type individuals generate transient mutant subpopulations or ``bubbles".
These bubbles are ordinarily doomed to extinction due to drift, but additional mutations in a lucky bubble may generate a complex adaptation that sweeps to fixation; this process has been referred to as ``tunneling" \citep{iwasa_2004, weissman_2009}.
On its own, valley crossing becomes more likely as the population size increases \citep{weissman_2009}. However, recent work suggests that valley crossing relative to the fixation of a simple beneficial mutation is least likely at intermediate population sizes where tunneling occurs \citet{ochs_2015}. This suggests that tunneling may be the most difficult mode of valley crossing. 
% Valley crossing in sexually reproducing populations was by \citet{weissman_2010}, which found that the criterion $r < s$, with $r$ the recombination rate and $s$ the selective advantage of the complex adaptation, is the necessary criterion for recombination to be helpful in crossing a valley.
% Fairly low recombination rates can bring mutant subpopulations together, increasing the rate at which complex adaptations arise and sweep.
% High recombination rates, on the other hand, lower the valley crossing rate by forcing fit individuals to outcross with the less fit wild type.

@@

\begin{figure}
\begin{center}
  \includegraphics[scale=0.75]{Figures/valley.pdf}
  \caption{Fitness landscape.}
  \label{fig:landscape}
\end{center}
\end{figure}

Our model features a population of $N$ haploid individuals in which an effectively infinite number of ``background" loci, with weak fitness effects, are currently segregating.
We focus on two loci, which are initially fixed for alleles $a$ and $b$.
These alleles mutate to $A$ and $B$ at rates $\mu_A$ and $\mu_B$, respectively (we neglect back mutations).
Genotypes $ab$, $Ab$, $aB$, and $AB$ have fitnesses $1$, $1-\delta_A$, $1-\delta_B$, and $1+s$ respectively, with $\delta \geq 0$ (intermediate genotypes are deleterious) and $1/N < s < 1$ (double mutant $AB$ is strongly beneficial).
For the remainder of our analysis, we will assume that $\mu_A = \mu_B = \mu$ and $\delta_A = \delta_B = \delta$.
We further assume that the large number of loci of weak effect contribute to a constant background fitness variance $\sigma^2$ which, by Fisher's ``fundamental theorem" of natural selection \citep{Fisher:1930}, sets the rate $v$ at which the mean fitness advances to $\sigma^{2}$.
If $x$ is the background fitness of a lineage, then the distribution of background fitnesses $f(x)$ is assumed to be roughly Gaussian: $f(x) = \frac{1}{\sqrt{2\pi}\sigma} \exp (\frac{(x-\bar{x})^2}{2\sigma^2})$.
Given this background fitness variance, we are interested in the expected time $\mathbb{E}\left[ T\right]$ for the double mutant $AB$ (our so-called ``complex adaptation'') to arise and fix in the population.

In very large populations, enough double mutants are generated that the valley crossing time is dominated by the time required for a double mutant to sweep through the population.
At the other extreme, in small populations, the crossing time is dominated by the wait for a deleterious single mutant to fix: only thereafter does a successful double mutant appear.
We begin by zeroing in on the intermediate case: stochastic tunneling, the most interesting and mathematically demanding mode of valley crossing.
We will review existing theory for this case, discuss the difficulty in the mathematical analysis due to genetic draft, and then present results for valley crossing times using forward-time simulations across a range of population sizes that include the sequential fixation, tunneling, and deterministic fixation cases.

There are three components to the process of stochastic tunneling.
First, a single mutant lineage must appear, which occurs with rate $N\mu$.
If this lineage arises at time $t_0$ and persists until time $T$, then it gives rise to $W = int_{t_0}^T n(t) dt$ total individuals before going extinct, where $n(t)$ is the number of mutant individuals extant at time $t$.
We refer to such a short-lived mutant lineage, and all the individuals generated within that lineage, as a mutant ``bubble": the time integrated number of individuals $W$ is the ``weight" of the bubble.
During the lifetime of a bubble, the complex adaptation must appear and establish; that is, it must rise to a high enough frequency that fixation is almost guaranteed, which occurs at rate $\mu \pfix$ where $\pfix$ is the probability that the double mutant reaches fixation from a single initial individual.
If it establishes, the complex adaptation will quickly sweep to fixation.
The first two steps can be folded in together and considered as one process, so that what is relevant is the expected time until a single mutant lineage arises that is destined to produce a successful double mutant.
In this way, $\mathbb{E}\left[ T \right] = \mathbb{E} \left[ T_0 + T_1 \right]$, where $T_0$ is the time to the first successful bubble--the first that produces a double mutant that is destined for fixation--and $T_1$ is the time required for the double mutant to sweep.
We assume that $T_0 \gg T_1$, so that the wait time is dominated by the first term
\comment{JVC: justify this briefly}.

Most of the time, a bubble will simply arise and go extinct.
If mutation and fixation are sufficiently rare events, then fixation of the double mutant can be modeled as a Poisson process: with probability $1-e^{-\mu \pfix W}$, the bubble with total size $W$ will give rise to a successful double mutant.
Computing the expected value of this quantity over all bubble sizes, $\phi = \left< 1-e^{\mu \pfix W} \right>$, gives the probability that some bubble will lead to a successful valley crossing.
One difficulty in computing the rate of tunneling in the high $N\sigma$ case where genetic draft is operating is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
That is to say, it is not possible to simply compute $\pfix$ and the weight distribution by themselves, but rather one must convolute both of them over the background fitness.
Individuals whose background fitness is near the nose of the fitness distribution will tend to give rise to longer-lived lineages, and if those individuals carry the complex adaptation, it will therefore likewise be more likely to fix.
Thus, $W$ will depend on the background fitness $x$ and the valley depth $\delta$, and $p_{\mathrm fix}$ will depend on $x$ and the peak height $s$.
These will have to be integrated over the fitness distribution $f(x)$ in order to obtain the total crossing probability $\Phi(s,\delta) = \int_{-\infty}^\infty \phi(x,s,\delta) f(x) dx$.
The expected valley crossing time then becomes $1/N\mu\Phi$.

\comment[inline]{
  \begin{itemize}      
  \item  This section is still too technical. For example, the reader here doesn't need to know what  ``weak driver'' is, or if they do, it should be more fully explained and used more than just in what is effectively a background section. 
  \item I think having the PDE for the fixation probability is ok, but you don't need the details about the limits necessary for deriving it. The point of keeping it more informational for the reader. 
  \item Also, avoid introducing new variables, like $D$, and keep the existing notation, like $\pfix$, instead of changing to $p$ since the simplicity doesn't save much time (we only use the equation once) and the extra notation could be confusing. 
  \item Part of the motivation for introducing the PDE in the first place is to then say what the results of analyzing it, namely what the fixation probability is under draft with fitness background $x$. You explain this a little bit, but you can more fully describe how the $\pfix$ increases exponentially then linearly. 
        \item Also, I don't think you mention how the valley depth $\delta$ and peak height $s$ related to the background fitness $x$ and fitness variance $\sigma$. This should probably come up here, but also probably in the first paragraph of the ``Mathematical background''
        \end{itemize}
      }
{\color{red}
We begin by considering the dynamics of $\pfix(x,s)$ before moving on to consider the distribution of bubble sizes.
We zero in on the ``weak driver" model, in which background mutations are frequent but of weak effect compared to $\sigma$.
In this case, the background beneficial mutation rate $U_b$ and mean fitness effect $\bar{\epsilon}$ turn out not to matter: what is relevant is the product $D = U_b \bar{\epsilon}^2/2$.
Taking the limit $U_b \to \infty$ and $\bar{\epsilon} \to 0$ yields the ``infinitesimal" model.
\citet{hallatschek_2011} and \citet{good_2014} showed that the fixation probability in this regime obeys the differential equation 
\begin{equation}
v\partial_x p(x,s) = x p(x,s) + D\partial_x^2 p(x,s) - \frac{p(x,s)^2}{2},
\end{equation}
in which we have assumed that $v \approx \sigma^2$, which is tantamount to claiming that mutational effects are less important than selection in determining the advance of the mean fitness.
The fixation probability then increases linearly in $x$ above some critical value $x_c$ and remains small otherwise, with $x_c \approx \sigma^4/4D$: when $\sigma$ is large, only mutations in the nose stand any chance of fixing unless they can be dragged to higher fitness by a background mutations (large $D$).
}

In order to compute the total valley crossing probability $\Phi$, the only remaining piece to compute is distribution of mutant bubble sizes $W$ as function of their background fitness $x$.
Since the tunneling probability for a lineage with background fitness $x$ is given by $\phi = \left< 1 - e^{-\mu p_{\mathrm fix} W} \right>$, we do not need to compute the full distribution of $W$ but rather only its Laplace transform: $\mathcal{L}\left[ p(w) \right] = \int_0^\infty e^{-zw} p(w) dw$, with $z = \mu \pfix$.
\citet{neher_shraiman_2011} showed that $\phi$ follows the equation
\begin{equation}
\partial_t \phi = z + (z + x - \sigma^2 t + \delta)\phi^2 - (1 + x - \sigma^2 t + \delta)\phi^2,
\end{equation}
but their analysis also included a recombination term that made it possible to pull out the integrated Laplace transform $\int_{-\infty}^\infty p(x) \phi(x) dx$ without the need to actually solve this equation or send time to infinity.
Unfortunately this method is not available to us, and the nose dominance of the bubble size distribution, combined with the $x$ dependence of $z$ (via $z = \mu \pfix(x)$), lead to difficulty in deriving a simple closed form expression for $\Phi$.
Therefor, we focus on simulation methods to study the dynamics of valley crossing.

\section*{Simulation methods}

@@

\begin{figure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/det_tunnel.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{Figures/det_fix.pdf}
\end{subfigure}
\caption{Valley crossing times in the semi-deterministic tunneling (left) and deterministic fixation (right) regimes. In both figures, $s = 10^{-2}$ and $\delta = 10^{-3}$: on the left, $N = 10^5$ and $\mu = 10^{-4}$, and on the right, $N = 10^6$ and $\mu = 10^{-3}$. Orange lines are averages, boxes are interquartile ranges, and end caps are overall ranges. Each box represents about $100$ simulation runs.}
\label{fig:deterministic}
\end{figure}

@@

Finally, we examine the ratio $\alpha = \tau_{\mathrm{valley}}/\tau_{\mathrm{sweep}}$, where $\tau_{\mathrm{sweep}}$ is the time required for a beneficial mutation to sweep to fixation and $\tau_{\mathrm{valley}}$ is the time required for a valley crossing event to occur.
By comparing how $\alpha$ changes as a function of the strength of selection on linked variation, $\sigma$, we can determine how $\sigma$ affects the speed of valley crossing relative to simple beneficial sweeps. Higher (or lower) values of $\alpha$ mean that valley crossing times increase (or decrease) relative to the time required for a simple beneficial sweep.
Though the precise value of this ratio depends on $\delta$ and the selection coefficients $s_{\mathrm{valley}}$ of the complex adaptation and $s_{\mathrm{sweep}}$ of the sweeping solo beneficial mutation, a general trend nonetheless emerges.
In this case, we set $s_{\mathrm{sweep}} = 0.01$, $s_{\mathrm{sweep}} = 0.1$, and $\mu = 10^{-5}$.
For several combinations of parameter values, we find that the ratio $\alpha$ is smaller for high $\sigma$ than for low $\sigma$: see figure \ref{fig:ratios}.
In figure \ref{fig:alpha_ratio}, we consider the behavior of $\alpha(10^{-2})/\alpha(10^{-6})$, a measure of the extent to which rapid adaptation speeds up valley crossing relative to sweeping (lower values indicate that rapid adaptation favors valley crossing).
Overall, we find that increasing $\sigma$ increases the extent to which valley crossing is favored \emph{except} when $N\mu > 1$.
In this case, stochastic tunneling ceases to be the dominant mode of valley crossing.
\citet{weissman_2009} referred to this regime as ``semi-deterministic tunneling": a more or less stable number of intermediate mutant individuals $N\mu/\delta$ persists in the population at low frequencies, which give rise to double mutants at a low but constant rate.
At high $\sigma$, the number of intermediate mutants is likely to fluctuate substantially \citep{cvijovic_2017}, but the average behavior is more or less the same.
This relation more or less obtains in a rapidly adapting population as well, with substantial fluctuations, so $\pfix$ becomes more important in determining the valley crossing rate: since this scales more weakly with $s$ under draft, the overall crossing rate is slowed.

@@

% One difficulty in computing the rate of tunneling in the high $N\sigma$ case is that the expected bubble size and fixation probability will both turn out to depend on the background fitness of an individual.
% That is to say, the total rate of valley crossing will be given by $\int P(x) \pfix(x,s) \mathbb{E}\left[ w(x,\delta) \right] dx$, with $P(x)$ the (presumably Gaussian) background fitness distribution.
% One cannot simply compute $\pfix$ and the weight distribution by themselves but rather must convolute both of them over the background fitness.
% This is intuitive: individuals near the nose of the distribution will tend to give rise to longer-lived lineages, and because they are near the nose, the complex adaptation will therefore likewise be more likely to fix.
% We proceed to derive both $\pfix(x,s)$ and $P(w)$ in the zero recombination limit.
